Estimating carrying capacity for juvenile salmon using quantile random forest models

### 2020-12-04 analysis/paper/paper.Rmd
added "Freshwater Ecology" as subtitle, to indicate what section it's being submitted under
removed some white space

---
title: Estimating carrying capacity for juvenile salmon using quantile random forest models
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: false
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      keep_tex: true
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |

  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids (*Oncorhynchus* spp.). A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest model to  provide estimates of habitat carrying capacity for Chinook salmon (*O. tshawytscha*) parr during the summer months, at both the site and watershed scale. Quantile random forest models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on quantile random forest models, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
  **Key words:** carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
keywords: |
  carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
highlights: |
  These are the highlights. 
---

>>>>

---
title: Estimating carrying capacity for juvenile salmon using quantile random forest models
subtitle: Freshwater Ecology
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: false
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      keep_tex: true
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |
  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids (*Oncorhynchus* spp.). A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest model to  provide estimates of habitat carrying capacity for Chinook salmon (*O. tshawytscha*) parr during the summer months, at both the site and watershed scale. Quantile random forest models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on quantile random forest models, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
  **Key words:** carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
keywords: |
  carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
highlights: |
  These are the highlights. 
---
